# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

Load the data.  The dataset looks easy to work with and will not be processed further.
```{r}
setwd("/home/bmarx/docs/coursera/reproducibleresearch/project1/RepData_PeerAssessment1")
unzip("activity.zip")
data <- read.csv("activity.csv",header=TRUE)
```

## What is mean total number of steps taken per day?

Make a histogram of the total number of steps taken each day.
```{r fig.width=12}
library(ggplot2)
p <- qplot(data$date,data$steps,geom=c("line")) + labs(title="Total Number of Steps Taken Each Day")
p + theme(axis.text.x = element_text(angle=90))
```

Calculate and report the mean total number of steps taken per day.
```{r}
mean <- aggregate(data$steps,list(data$date),FUN="mean")
names(mean) <- c("Date","Mean")
mean
```

Calculate and report the median total number of steps taken per day.
```{r}
median <- aggregate(data$steps,list(data$date),FUN="median")
names(median) <- c("Date","Median")
median
```


## What is the average daily activity pattern?

Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
daily_pattern <- aggregate(data$steps ~ data$interval,list(data$date),FUN="mean")
names(daily_pattern) <- c("interval","mean")
plot(daily_pattern,type="l",main="Average Steps Taken Per Interval Over All Days")
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
daily_pattern[daily_pattern$mean == max(daily_pattern$mean),]
```


## Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
nrow(data)-sum(complete.cases(data))
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

The mean value for the five minute interval over all days will be used.  This value was already calculated and is stored in daily_pattern.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
filled_in_dataset <- data
for (i in 1:nrow(filled_in_dataset)){
     if(is.na(filled_in_dataset[i,1])){
          filled_in_dataset[i,1] = daily_pattern[daily_pattern$interval == filled_in_dataset[i,3],2]
     }
}
```

Make a histogram of the total number of steps taken each day?
```{r fig.width=12}
library(ggplot2)
p <- qplot(filled_in_dataset$date,filled_in_dataset$steps,geom=c("line")) + labs(title="Total Number of Steps Taken Each Day w/ Missing Values Added In")
p + theme(axis.text.x = element_text(angle=90))
```

Calculate and report the mean total number of steps taken per day.
```{r}
filled_in_mean <- aggregate(filled_in_dataset$steps,list(filled_in_dataset$date),FUN="mean")
names(filled_in_mean) <- c("Date","Mean")
filled_in_mean
```

Calculate and report the median total number of steps taken per day.
```{r}
filled_in_median <- aggregate(filled_in_dataset$steps,list(filled_in_dataset$date),FUN="median")
names(filled_in_median) <- c("Date","Median")
filled_in_median
```

Do these values differ from the estimates from the first part of the assignment? 
```{r fig.width = 12}
diff <- c()
for (i in 1:nrow(mean)){
     if(is.na(mean[i,2])){
          diff <- rbind(diff,filled_in_mean[i,2]-0)
     }
     else{
          diff <- rbind(diff,filled_in_mean[i,2]-mean[i,2])
     }
}
diff <- cbind(mean,diff)
names(diff) <- c("Date","Mean","d")
p <- qplot(diff$Date,diff$d,geom=c("point")) + labs(title="Difference Between Average Steps Taken Per Day of Original and Filled in Datasets")
p + theme(axis.text.x = element_text(angle=90))
```

What is the impact of imputing missing data on the estimates of the total daily number of steps?

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
library(lattice)
data$date <- as.Date(data$date)
day_type <- c()
for (i in 1:nrow(data)){
     if(weekdays(data[i,2]) == "Saturday" || weekdays(data[i,2]) == "Sunday"){
          day_type = rbind(day_type,"Weekend")
     }
     else{
          day_type = rbind(day_type,"Weekday")
     }
}
data <- cbind(data,day_type)
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
weekdays <- data[data$day_type == "Weekday",]
weekends <- data[data$day_type == "Weekend",]

weekday_avg <- aggregate(weekdays$steps ~ weekdays$interval,list(weekdays$date),FUN="mean")
names(weekday_avg) <- c("interval","mean")

weekend_avg <- aggregate(weekends$steps ~ weekends$interval,list(weekends$date),FUN="mean")
names(weekend_avg) <- c("interval","mean")

weekday_avg <- cbind(weekday_avg,rep("Weekday",nrow(weekday_avg)))
names(weekday_avg) <- c("interval","mean","day_type")
weekend_avg <- cbind(weekend_avg,rep("Weekend",nrow(weekend_avg)))
names(weekend_avg) <- c("interval","mean","day_type")

combined <- rbind(weekday_avg,weekend_avg)
xyplot(mean~interval|day_type,data=combined,layout=c(1,2),type="l")
```


